---
title: "API 222 Problem Set 2"
subtitle: "Machine Learning and Big Data Analytics: Fall 2024"
output: pdf_document
date: "Due at 11:59am on October 23 - submit on Gradescope"
---

```{r, include=FALSE, message=FALSE, warning=FALSE}
# The needed libraries
library(tidyverse)
library(stargazer)
library(FNN)
library(kableExtra)
library(glmnet)

# Ask R to not present numbers using scientific notation
options(scipen = 999)
```

This problem set is worth 30 points in total. To get full credit, submit your code along with a write-up of your answers. This should either be done in R Markdown or Jupyter Notebook, submitted in one knitted PDF.


## Final Project Groups (0 pts)

Please join one of 30 final project groups that have been created on this \color{blue} [Canvas](https://canvas.harvard.edu/courses/146422/groups#tab-26761) \color{black} page. 

Details about the final project can be found on this \color{blue} [Canvas](https://canvas.harvard.edu/courses/146422/files/folder/Final%20Project) \color{black} page. You just need to form your group by October 23. The main project milestones will be after the midterm. We recommend forming groups of 5 students. All students working together should join the same group. PhD students need to work individually (see details on Canvas). If you are a PhD student or are otherwise working alone, please form a group by yourself. Please email Jacob if you have questions about this assignment or the final project.

## Conceptual Questions (15 pts)

1. Consider the four main classification methods that have been presented thus far this semester: logistic regression, k-Nearest Neighbors, linear discriminant analysis (LDA), and quadratic discriminant analysis (QDA. Which of these methods may be appropriate if you know the decision boundary between the classes is linear? (3pts)



2. Suppose you had the following data and you are using KNN Regression with Euclidean distance. Consider the prediction problem where you want to predict Y for the data point X1 = X2 = X3 = 0.

| X1 | X2 | X3 | Y |
|----|----|----|---|
| 0  | 3.5  | 2  | 2 |
| 1  | 2.1  | 3  | 1 |
| 2  | 4.7  | 1  | 3 |
| 1  | 3.9  | 1  | 2 |
| 0  | 2.9  | 2  | 4 |
| 1  | 1.5  | 2  | 1 |
| 1  | 3.5  | 4  | 2 |


(a) Compute the Euclidean distance between each observation and the test point, X1 = X2 = X3 = 0. (1pt)


(b) What is your prediction with K = 2? Why? (1pt)


(c) If the Bayes decision boundary in this problem is highly nonlinear, then would we expect the best value for K to be large or small? Why? (1pt)


3. Consider we conduct a research study analyzing the risk factors for developing prostate cancer among men, with variables \(X_1 = \text{age (years)}\), \(X_2 = \text{family history of prostate cancer (0 = no, 1 = yes)}\), \(X_3 = \text{smoking status (0 = non-smoker, 1 = smoker)}\), and \(Y = \text{probability of developing prostate cancer}\). A logistic regression analysis is performed, resulting in estimated coefficients \(\hat{\beta}_1 = 0.06\), \(\hat{\beta}_2 = 1.2\), \(\hat{\beta}_3 = 0.8\), and \(\hat{\beta}_0 = -3.5\).

(a) Interpret \(\hat{\beta}_2\). (1 pt)




(b) Estimate the probability that a 60-year-old man with a family history of prostate cancer who is a smoker develops prostate cancer. (2 pts)




4. k-fold cross-validation

(a) Briefly explain how k-fold cross-validation is implemented. (2pts)



(b) What are the advantages of k-fold cross-validation relative to the validation set approach? (1pt)



5. Suppose you want to minimize the false negative rate in your classification. You run two models: A and B. AUC for Model A is 0.7 and for Model B is 0.8. Can you conclude that you should choose Model B? Why or why not? (3 pts)

\newpage



## Applied Questions (15 pts)

**Predicting Hospital Length of Stay**

For the next portion of this assignment you will be working with the `LengthOfStay.csv` dataset. This dataset has data points on patients admitted into hospital, indicators of their health condition and how long they were admitted in the hospital. 

This is an important problem in healthcare. In order for hospitals to optimize resource allocation, it is important to predict accurately how long a newly admitted patient will stay in the hospital. 

1. What are the dimensions of the dataset? (1 pt)

```{r, message=FALSE, warning=FALSE}
# code
```

2. Use the `cor()` function to display the correlations of all **continuous** variables in the dataset. Which variables is most highly correlated with `lengthofstay`? (2 pts)

```{r}
# code
```




Consider the prediction problem where you want to predict the length of stay for a patient (`lengthofstay`) against all other variables available in the data set.

3. Run ridge regression with cross-validation and standardized features using the canned function `cv.glmnet` from the package `glmnet`. You can use the $\lambda$ sequence generated by `cv.glment` (you do not need to provide your own $\lambda$ sequence). In order to receive credit for this question, make the line immediately preceding this command say `set.seed(222)` and run the two lines together. Please report all numbers by rounding to three decimal places. (2 pts)

```{r}
# code
```


(a) Which $\lambda$ had the lowest mean cross-validation error? (1 pt)


```{r}
# code
```


(b) What was the cross-validation error? (1 pt)

```{r}
# code
```

(c) What was the standard error of the mean cross-validation error for this value of $\lambda$? (1 pt)

```{r}
# code
```


(d) What was the largest value of $\lambda$ whose mean cross validation error was within one standard deviation of the lowest cross-validation error? (1 pt)

```{r}
# code
```

4. Produce the regression coefficients for the ridge regression model with the $\lambda$ value that minimizes the cross-validation error. Compare these coefficients with a standard linear regression model. (2 pts)

```{r}
# code
```

5. Now consider the same prediction problem. Implement your own 5-fold cross-validation routine for KNN for $K = 1, ..., 50$ (write the cross-validation routine yourself rather than using a canned package). Include the snippet of code you wrote here. It should not exceed 20 lines. (4pts)

```{r, message=FALSE, warning=FALSE}
# code
```

(a) Plot of mean cross-validation MSE as a function of $k$. 

```{r}
# code
```


(b) The best k according to CV is


```{r}
# code
```



(c) The cross-validation error for the best k is


```{r}
# code
```



\newpage


## Challenge Problem (just for fun 0 pts)

Tasks:

(a) Predict Length of Stay with LASSO Regression with Cross-Validation:

- Use the `cv.glmnet` function from the `glmnet` package to perform LASSO regression with cross-validation.

- Set `alpha = 1` to specify LASSO regression.

(b) Selecting the Best Model:

- Identify the value of lambda that minimizes the cross-validation error.

- Report this value of lambda and the corresponding cross-validation error (rounded to three decimal places).

(c) Visualization:

- Plot the cross-validation curve (cross-validation error versus log(lambda)) using `plot(cv.glmnet_object)`.

- Plot the coefficient paths as a function of log(lambda) to visualize how coefficients change with different lambda values.

(d) Interpretation:

- Discuss the variables that have been selected by the LASSO model.

- Compare these variables to those in the standard linear regression model you previously fitted.

- Comment on any differences and provide possible explanations for why certain variables were eliminated.

```{r}
# code
```


